# Example session to demonstrate R's capabilities

R can behave just like any calculator - it understands numbers!

```{r}
1

1 + 1

(1 + 5) / 3
# now try some of your own calculations
```

R also allows collections of numbers to be operated on at once.
Collections of text and numbers can be assigned names - these are *objects*.

```{r}
c(1, 2, 3) / 2 # note that each number is divided by 2 - it knows what you mean

x <- c(1, 2, 3) # assign the numbers to an object
x * 5
x^3
x * x

y <- c(4, 2, 1)
plot(x, y) # very basic plotting
lines(x,y)
```

R is very versatile. It can do almost *anything* involving numbers and
text. However, it was designed to work with fairly large datasets. Let's load one:

```{r}
# navigate to data - modify this depending on your dataset:
df <- read.csv("~/repos/smsim-course/data/cakeMap/ind.csv") 
head(df) # this asks R to print the top rows of the file
summary(df$Car) # note the use of the $ symbol to specify columns
summary(df$ageband4) # note the output is different - this is because of the type of data:
class(df$Car) # this was recognised by R as numeric integer data
class(df$ageband4) # this is factor data
```

Let's explore the data a little - can we identify a link between cake consumption and class?

```{r}
table(df$NCakes, df$NSSEC8)
```

Perhaps, but let's recategorise the cake data so it's numeric:

```{r}
library(car) # load a new package - try install.packages("car") if this fails
levels(df$NCakes) <- c(0.5, 1.5, 4, 10, 0.1)
df$NCakes <- as.numeric(as.character(df$NCakes))
class(df$NCakes)
class(df$NSSEC8)
df$NSSEC8 <- paste0("ns",df$NSSEC8)
aggregate(df$NCakes ~ df$NSSEC8, mean, data=df)
```

How does cake consumption vary by age? Again, R can tell us:

```{r}
aggregate(df$NCakes ~ df$ageband4, mean, data=df)
```


This ability to use numbers in a general purpose way and
handle large datasets makes R extremely flexible.
The flexibility also means R is hard - this is why we need tutorials
like this one to see how it works.

